---
title: "CODEX data analysis: Unsupervised clustering"
output:
  html_document:
    df_print: paged
---

## load python env
```{r}
# Requires python virtual environment for the leiden package
reticulate::use_virtualenv("venv_scimap/", required=TRUE)
```

## Load libraries and functions

```{r,message=F,warning=F}
# remotes::install_github("satijalab/seurat", "feat/imaging")
# remotes::install_github("mojaveazure/seurat-disk")
library(Seurat)
library(ggplot2)
library(reticulate)
library(viridis)
library(dplyr)
library(stringi)
library(ClusterR)
library(uwot)
library(patchwork)
source("utils/R_utils.R")

```

## Load aggregated and cleaned object
```{r,cache=F}
sobj<- readRDS('Path_to_saved_object.rds')
```

## Number of threads
```{r}
n_threads <- 10
```

## Define structural markers
```{r,cache=F}
struc_markers <- c('CD3e',"CD4","CD8","CD56","DC-LAMP","HLA-DR","Podoplanin",'CD20','CD31','CD68',"CD163",'CD66b',
                  'Pancytokeratin',"E-Cadherin","Vimentin","N-Cadherin","SLUG","FAP","Blank-1")
struc_markers <- paste0(struc_markers,"-nuc")
```


## Perform Kmeans and subset for low dimension umap
```{r,eval=T,cache=F}
sobj_counts<-as.matrix(t(sobj@assays$zscores@counts[struc_markers,]))
kmc_emb<- MiniBatchKmeans(data = sobj_counts, # now performing kmeans on struct markers as well
                      batch_size = 100000, # making sure to get cells of interest
                      clusters = 200,
                      max_iters = 150,
                      seed = 8
                        )
kmc <- predict(kmc_emb,newdata = sobj_counts)
sobj <- AddMetaData(sobj,col.name = "kmc",metadata = factor(kmc))
Idents(object = sobj) <- factor(kmc)
sobj_small <- subset(sobj, downsample = 3000,idents =unique(kmc)) #subset 600k cells

```

```{r,eval=T,cache=F}
train_df <- data.frame(t(sobj_small@assays$zscores@counts[struc_markers,]))
```

```{r}
k=30
```
## Run initial umap

These next steps are dedicated to finding best values of a and b for the umap, they are therefore set as eval=F once a and b have been found
```{r,eval=F}
# t-UMAP is equivalent to a = 1, b = 1
# remember to get the nearest neighbor data back too
k=30
init_a1b1 <- tumap(train_df, ret_nn = TRUE,metric = "cosine",n_neighbors = k,batch = TRUE,fast_sgd=T,n_sgd_threads = n_threads,n_threads = n_threads)
```

## Determine best umap parameters
```{r,fig.height=7,fig.width=10,eval=F}
# find a
b<-0.7
a_vals <- c(5,3,2,1)
a_res <- lapply(a_vals,function(a){
  print(a)
  train_umap_2D <- umap(train_df, nn_method = init_a1b1$nn, a = a, b = b,
                   batch = TRUE,
                   fast_sgd=T,
                   n_neighbors = k, #50 was ok
                   metric = "cosine",
                   n_sgd_threads = n_threads,
                   n_threads = n_threads)
  embeddings_train_2D <- as.matrix(train_umap_2D)
  colnames(embeddings_train_2D) <- paste0("UMAP_", 1:2)
  sobj_small@reductions[["umap"]] <- CreateDimReducObject(embeddings = embeddings_train_2D, key = "UMAP_", assay = DefaultAssay(sobj_small))
  # Quick look at the umap
  pl <- DimPlot(sobj_small,reduction = 'umap',group.by = 'orig.ident')+ ggtitle(paste0("a=",a," b=",b)) + NoLegend() + NoAxes() + coord_fixed()
  pl[[1]]$layers[[1]]$aes_params$alpha <- .5
  pl
})
# find b based on whichever value of a you prefer
# a = 3 is not bad
a <- 3
b_vals <- seq(from=0.5,to=0.8,by=0.05)
b_res <- lapply(b_vals,function(b){
  print(b)
  train_umap_2D <- umap(train_df, nn_method = init_a1b1$nn, a = a, b = b,
                   batch = TRUE,
                   fast_sgd=T,n_neighbors = k,
                   metric = "cosine",
                   n_sgd_threads = n_threads,
                   n_threads = n_threads)
  embeddings_train_2D <- as.matrix(train_umap_2D)
  colnames(embeddings_train_2D) <- paste0("UMAP_", 1:2)
  sobj_small@reductions[["umap"]] <- CreateDimReducObject(embeddings = embeddings_train_2D, key = "UMAP_", assay = DefaultAssay(sobj_small))
  # Quick look at the umap
  pl <- DimPlot(sobj_small,reduction = 'umap',group.by = 'orig.ident')+ ggtitle(paste0("a=",a," b=",b)) + NoLegend() + NoAxes() + coord_fixed()
  pl[[1]]$layers[[1]]$aes_params$alpha <- .5
  pl
})
a_res <- lapply(a_res,function(x){
  x$layers[[1]]$geom_params$pointsize <- 1
  x
})
b_res <- lapply(b_res,function(x){
  x$layers[[1]]$geom_params$pointsize <- 1
  x
})
wrap_plots(a_res,ncol=4)
wrap_plots(b_res,ncol=4)
```

## Learn Umap with identified parameters (need to re-learn it to get the model)
```{r,cache=F}
set.seed(8)
train_umap_2D <- umap(train_df,
                   batch = TRUE,
                   fast_sgd=T,
                   n_sgd_threads = n_threads,
                   n_threads = n_threads,
                   ret_model = TRUE,
                   ret_extra = c("nn",'fgraph'),
                   n_neighbors =k,# check option to set off for clustering
                   verbose = T,
                   n_components = 2,
                   a=3,b=0.6, # a=3,b=0.6 give more spread out results
                   metric = 'cosine')
#Add 2D umap to subset seurat object
embeddings_train_2D <- as.matrix(train_umap_2D$embedding)
colnames(embeddings_train_2D) <- paste0("UMAP_", 1:2)
sobj_small@reductions[["umap"]] <- CreateDimReducObject(embeddings = embeddings_train_2D, key = "UMAP_", assay = DefaultAssay(sobj_small))
#Quick look at the umap
pl <- DimPlot(sobj_small,reduction = 'umap',group.by = 'orig.ident',pt.size = 1,shuffle = T)
pl[[1]]$layers[[1]]$aes_params$alpha <- .5
pl
```
## clean memory before clustering
```{r}
gc()
Sys.sleep(5)
```

## Run Leiden clsutering
```{r,fig.height=8,fig.width=8,cache=F}

# leiden modularity partition on fgraph
options(scipen = 1000000000)
t0 <- Sys.time()
leiden_modul <- suppressMessages(suppressWarnings(leiden::leiden(train_umap_2D$fgraph,
                                                         resolution_parameter=1.8,
                                                         seed = 1,
                                                         partition_type ="ModularityVertexPartition")))
t1 <- Sys.time()
print(round(t1-t0))
sobj_small<- AddMetaData(sobj_small,factor(leiden_modul),col.name = "leiden_modularity")
leiden_pl <- DimPlot(sobj_small,reduction = 'umap',group.by = 'leiden_modularity',label = T,label.box = T,repel=T,pt.size=1) + NoLegend() + ggtitle(paste0("leiden Modularity partition, time= ",round(t1-t0),"min")) + coord_fixed()
sobj_small$leiden <-  sobj_small$leiden_modularity
leiden_pl
```

## Refine leiden clusters
```{r,fig.height=10,fig.width=20,eval=T,cache=F}
sobj_small <- refine_labels(sobj_small,label_name = "leiden",k_neighbors = 100,n_threads = n_threads,Kmean_k = 3)

plot_refined <- DimPlot(sobj_small,reduction = 'umap',group.by = 'leiden_refined',label = T,label.box = T,repel=T,pt.size=1) + NoLegend() + coord_fixed()
plot_unrefined <- DimPlot(sobj_small,reduction = 'umap',group.by = 'leiden',label = T,label.box = T,repel=T,pt.size=1) + NoLegend() + coord_fixed()
plot_unrefined + plot_refined
```

## Show expression of markers (zscores)
```{r,message=F,warning=F,fig.width=20,fig.height=15,dpi=300,eval=T,cache=F}
DefaultAssay(sobj_small) <- "zscores"
pl <- FeaturePlot(sobj_small,
            reduction = "umap",
            features = struc_markers,
            pt.size=1,
            ncol=6) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brightest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 5),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
pl <- FeaturePlot(sobj_small,
            reduction = "umap",
            features = c("DC-LAMP-nuc","Podoplanin-nuc","HLA-DR-nuc","MHC-1-nuc"),
            pt.size=1,
            #cols = viridis(100,option = "turbo"),
            ncol=2) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brighest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 5),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
```

## Show expression of markers (MFI)
--> adjust limits according to your data dynamic range
```{r,message=F,warning=F,fig.width=20,fig.height=15,dpi=300,eval=T,cache=F}
DefaultAssay(sobj_small) <- "MFI"
pl <- FeaturePlot(sobj_small,
            reduction = "umap",
            features = struc_markers,
            pt.size=1,
            ncol=6) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brightest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 250),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
pl <- FeaturePlot(sobj_small,
            reduction = "umap",
            features = c("DC-LAMP-nuc","Podoplanin-nuc","HLA-DR-nuc","MHC-1-nuc"),
            pt.size=1,
            #cols = viridis(100,option = "turbo"),
            ncol=2) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brighest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 250),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
```

## Predict umap on the whole dataset
```{r,fig.height=10,fig.width=10,cache=F}
#Predict on the full dataset 
test_df <- sobj@assays$zscores@counts[struc_markers,]
test_umap_2D <- umap_transform(data.frame(t(test_df)),
                               model =train_umap_2D,
                               batch = T,
                               n_sgd_threads = n_threads,
                               ret_extra = c("nn"),
                               n_threads = n_threads,
                               verbose = TRUE)
# Add 2D umap to full seurat object 
colnames(test_umap_2D$embedding) <- paste0("UMAP_", 1:2)
sobj@reductions[["umap"]] <- CreateDimReducObject(embeddings = test_umap_2D$embedding, key = "UMAP_", assay = DefaultAssay(sobj))
#Quick look at the umap
pl <- DimPlot(sobj,reduction = 'umap',group.by = 'orig.ident',raster.dpi = c(1080,1080),pt.size = 1 ) + coord_fixed()
pl[[1]]$layers[[1]]$aes_params$alpha <- .5
pl
```

## Predict leiden labels on whole dataset
```{r,fig.height=10,fig.width=10,cache=F}
test_labels <- predict_labels_nn(umap_test = test_umap_2D,train_labels = factor(sobj_small$leiden_refined),method = 'vote')
sobj<- AddMetaData(sobj,factor(test_labels),col.name = "leiden")
pl <- DimPlot(sobj,reduction = 'umap',group.by = 'leiden',label = T,label.box = T,repel=T,raster.dpi = c(1080,1080),pt.size=1) + NoLegend() + ggtitle(paste0("predicted leiden: vote")) + coord_fixed()
#pl[[1]]$layers[[1]]$aes_params$alpha <- .5
pl
Idents(sobj) = "leiden"
```

## Refine leiden clusters
```{r,fig.height=10,fig.width=20,eval=T,cache=F}
sobj <- refine_labels(sobj,label_name = "leiden",k_neighbors = 100,n_threads = n_threads,Kmean_k = 3)

plot_refined <- DimPlot(sobj,reduction = 'umap',group.by = 'leiden_refined',label = T,label.box = T,repel=T,raster.dpi = c(1080,1080),pt.size=1) + NoLegend() + coord_fixed()
plot_refined[[1]]$layers[[1]]$aes_params$alpha <- .5
plot_unrefined <- DimPlot(sobj,reduction = 'umap',group.by = 'leiden',label = T,label.box = T,repel=T,raster.dpi = c(1080,1080),pt.size=1) + NoLegend() + coord_fixed()
plot_unrefined[[1]]$layers[[1]]$aes_params$alpha <- .5

plot_unrefined + plot_refined
Idents(sobj) = "leiden"
```

## Visualize umap
```{r,fig.height=10,fig.width=20,cache=F}
#global umap with kmeans clusters
umap_clustering <- DimPlot(sobj,
        label = T,label.box = T,label.size = 7,
        group.by = 'leiden',
        raster.dpi = c(1080,1080),pt.size=1,
        reduction = 'umap') + ggtitle(paste0('n=',ncol(sobj),' cells')) + coord_fixed() + NoLegend()
#sample specific umap
umap_sample <- DimPlot(sobj,
                       reduction = 'umap',
                       raster.dpi = c(1080,1080),pt.size=1,
                       group.by = 'orig.ident') + coord_fixed()
umap_sample[[1]]$layers[[1]]$aes_params$alpha <- .7

umap_clustering+ umap_sample
```

## Visualize channels
```{r,message=F,warning=F,fig.width=20,fig.height=15,dpi=300,eval=T,cache=F}
DefaultAssay(sobj) <- "zscores"
pl <- FeaturePlot(sobj,
            reduction = "umap",
            features = struc_markers,
            pt.size=1,
            ncol=6) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brightest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 5),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
pl <- FeaturePlot(sobj,
            reduction = "umap",
            features = c("DC-LAMP-nuc","Podoplanin-nuc","HLA-DR-nuc","MHC-1-nuc"),
            pt.size=1,
            ncol=2) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brighest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 5),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
```
## Visualize channels MFI
```{r,message=F,warning=F,fig.width=20,fig.height=15,dpi=300,eval=T,cache=F}
DefaultAssay(sobj) <- "MFI"
pl <- FeaturePlot(sobj,
            reduction = "umap",
            features = struc_markers,
            raster.dpi = c(1080,1080),pt.size = 1,
            ncol=7) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brightest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 250),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
pl <- FeaturePlot(sobj,
            reduction = "umap",
            features = c("DC-LAMP-nuc","Podoplanin-nuc","HLA-DR-nuc","MHC-1-nuc","GZMB-nuc","GZMK-nuc"),
            raster.dpi = c(1080,1080),pt.size = 1,
            ncol=3) 
pl[[1]]$layers[[1]]$aes_params$alpha <- .5

# custom color scale, values below 0 are set to darkest color, values above 3 to the brighest color (red with this color palette)
pl & scale_color_gradientn(colours = viridis::viridis(100,option='viridis'),
                                limits = c(0, 250),
                                oob = scales::squish) & NoLegend() & coord_fixed() & NoAxes() &DarkTheme()
```
## Aggregate and annotate clusters
```{r,eval=F}

annot <- list(CD8_T_cells = c(10,6),
              CD4_T_cells = c(23,42,14),
              B_cells = c(38,24),
              Dendritic_cells = c(28),
              Macrophages_CD68 = c(21,22,18,39,15,19),
              Macrophages_CD163 = c(1),
              Neutrophils = c(20),
              Endothelial_cells = c(41),
              Fibroblasts = c(34,33,35),
              Lymphatics = c(2,40),
              Tumor = c(36,32,25,11,27,30,17,9,29,31,26,7,8,4,3,5),
              CD56_pop = c(16),
              Autofluo = c(44,12,13))
#find missing : 
sort(unlist(annot))
     
sobj$annotation <- "other"
for (i in names(annot)){
  sobj$annotation[sobj$leiden %in% unlist(annot[i])] <- i
}

sobj$annotation <- factor(sobj$annotation,levels=c("Neutrophils","Lymphatics","B_cells","Dendritic_cells","Macrophages_CD68","Macrophages_CD163","CD8_T_cells","CD4_T_cells","CD56_pop",
                                                   "Endothelial_cells","Fibroblasts","Tumor","Tumor_mesenchymal","Autofluo","other"))

```

```{r}
saveRDS(sobj,'Path_to_saved_object_2.rds')
```

## Visualize cell population annotations
```{r,fig.height=12,fig.width=12,eval=F}
library(pals)
clusters <- levels(sobj$annotation)
set.seed(4854)
pop_colors <- sample(as.character(pals::brewer.accent(length(clusters))))
pop_colors <- as.character(pals::brewer.paired(length(clusters)))

#change other to grey 
oid <- grep('other',clusters)
oifd <-  grep('Autofluo',clusters)
# pop_colors[oid] <- 'grey'
# pop_colors[oifd] <- 'azure3'

pl <- DimPlot(sobj,
        pt.size = 0.001,
        raster=F,
        repel = T,
        label = T,
        label.box = T,label.size = 7,
        label.color = 'black',
        group.by = 'annotation',
        reduction = 'umap') + scale_color_manual(values = pop_colors,
                       breaks = clusters) + coord_fixed() + NoLegend() + scale_fill_manual(values = pop_colors,breaks = clusters)
pl[[1]]$layers[[1]]$aes_params$alpha <- .4
pl
```

## Visualize sample composition
```{r,eval=F,fig.width=12,fig.height=6}
df <- data.frame(sobj$annotation)
colnames(df)<- 'annotation'
df$sample <- sobj$orig.ident

counts <- group_by(df, annotation, sample) %>% summarise(count = n())

# Stacked + percent
ggplot(counts, aes(fill=annotation, y=count, x=sample)) + 
  geom_bar(position="fill", stat="identity") +
  scale_fill_manual(values = pop_colors,
                    breaks = clusters)+
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1),axis.title.x = element_blank())
```
## Visualize umap split by sample
```{r,fig.height=20,fig.width=30,eval=F,dpi=300}
DimPlot(sobj,
        split.by = 'orig.ident',
        group.by = 'cell_population',
        pt.size = 0.01,
        #cols=colors,
        raster = F,
        ncol=8,
        reduction = "spatial_") &
  theme(legend.position="bottom") & 
  coord_fixed() &
  NoAxes() &
  scale_y_reverse() &
  scale_color_manual(values = pop_colors,
                     breaks = clusters)
```

## Visualize umap for specific sample
```{r,fig.height=20,fig.width=20,eval=F,dpi=500}

DimPlot(sobj,
        cells = names(which(sobj$orig.ident == 'P02_BL')),
        group.by = 'annotation',
        raster = F,
        reduction = "spatial_") &
  theme(legend.position="bottom") & 
  coord_fixed() &
  NoAxes() &
  scale_y_reverse() &
  scale_color_manual(values = pop_colors,
                     breaks = clusters)
DimPlot(sobj,
        cells = names(which(sobj$orig.ident == 'P02_C2')),
        group.by = 'annotation',
        raster = F,
        reduction = "spatial_") &
  theme(legend.position="bottom") & 
  coord_fixed() &
  NoAxes() &
  scale_y_reverse() &
  scale_color_manual(values = pop_colors,
                     breaks = clusters)
```

## Write h5ad for Tissuumaps
```{r}
library(anndata)

t1 <- Sys.time()
paths <- "tissuumap_folder"
for(x in unique(sobj$orig.ident)){
  meta_df <- sobj@meta.data[sobj$orig.ident == x ,]
  #x_mat <-  t(sobj@assays$zscores@counts[,sobj$orig.ident == x])
  ad <- anndata::AnnData(obs=meta_df,shape = dim(meta_df))
  path_to_save <- paste0(paths,x,"/seurat_tmap.h5ad")
  suppressMessages(anndata::write_h5ad(anndata = ad,filename = path_to_save))
}
t2 <- Sys.time()
t2-t1
```


