---
title: "CODEX data analysis: Spatial interactions"
output: html_notebook
---


## load python env
```{r}
# Necessary for scimap
reticulate::use_virtualenv("venv_scimap/", required=TRUE)
```

```{r,message=F,warning=F}
# remotes::install_github("satijalab/seurat", "feat/imaging")
# remotes::install_github("mojaveazure/seurat-disk")
library(Seurat)
library(ggplot2)
library(dplyr)
library(stringi)
library(uwot)
library(patchwork)
library(readxl)
library(stringr)
library(tidyr)
library(gridExtra)
library(data.table)
library(reticulate)
source("~/script/maxime.utils.r")
source("utils/R_utils.R")
```

```{r}
sobj<- readRDS('Path_to_saved_object_3.rds')
```

```{r}
sobj <- AddMetaData(sobj,sobj@reductions$umap@cell.embeddings[,1],"umap_1")
sobj <- AddMetaData(sobj,sobj@reductions$umap@cell.embeddings[,2],"umap_2")
meta_df <- sobj@meta.data
ad <- anndata::AnnData(obs=meta_df,shape = dim(meta_df))
```

this chunk has to be run first
```{python}
import numpy as np
print("Numpy needs to be imported in this chunk for some reason")
```

## python code for spatial interactions
```{python}
import anndata as ad
import pandas as pd
import scimap as sm

adata = r.ad
```
## Compute distance
```{python}
adata = sm.tl.spatial_distance (adata,
                               x_coordinate='x', y_coordinate='y',
                               z_coordinate=None,
                               phenotype='ann_v3',
                               subset=None,
                               imageid='orig.ident',
                               label='spatial_distance')
```

## compute spatial interactions
```{python}
adata = sm.tl.spatial_interaction (adata, 
                                  method='radius', 
                                  x_coordinate='x', y_coordinate='y', 
                                  z_coordinate=None, 
                                  phenotype='ann_v3', 
                                  imageid='orig.ident', 
                                  radius=30, 
                                  label='spatial_interaction_radius')
```

## Generate video from spatial coords to umap
```{python}
#sm.hl.animate (ad_with_obs, color='cell_population',embedding='X_umap',x_coordinate='x', y_coordinate='y',subset=['P02_BL'],figsize=(20,20),n_frames=500,final_frame=20,alpha=0.5,watermark=False,
#               save_animation = 'res/animate')
```

## Measure Pscore between CD4 T cells and CD8 T cells
```{python}
adata = sm.tl.spatial_pscore(adata,
                               proximity= ['CD4_T_cells','CD8_T_cells'],
                               score_by = 'orig.ident',
                               imageid='orig.ident', 
                               x_coordinate='x', y_coordinate='y', 
                               phenotype='annotation',
                               method='radius',
                               radius=30,
                               label='spatial_pscore')
```

Plot Pscores
```{python}
import matplotlib.pyplot as plt
plt.figure(figsize=(10, 5))
sm.pl.spatial_pscore (adata, color='Black', plot_score='Proximity Density')
plt.show()
```

## Get spatial interactions back into R
```{r}
py$adata
interactions <- as.data.frame(py$adata$uns$spatial_interaction_radius)
dists <-  as.data.frame(py$adata$uns$spatial_distance)

sobj@misc$interactions <- interactions
sobj@misc$dists <- dists
```

## Interactions between all samples
--> specfic to my project, could be useful for other projects
```{r,fig.height=15,fig.width=15}
library(pheatmap)
#extract spatial interactions from object

pheno_list <- setdiff(unique(interactions$phenotype),c("all_neg","Autofluo","other"))
BL <- unique(sobj$orig.ident[sobj$timepoint_s == "BL"])

int_avg <- sapply(pheno_list,function(pop){
  int_df <- interactions[interactions$phenotype==pop,c("phenotype","neighbour_phenotype",BL)]
  int_df <- int_df[match(pheno_list,int_df$neighbour_phenotype,),BL]
  rowMeans(int_df,na.rm = T)
})
rownames(int_avg) <- pheno_list

#summarize spatial interactions for all images
int_pval <- sapply(pheno_list,function(pop){
  int_p <- interactions[interactions$phenotype==pop,c("phenotype","neighbour_phenotype",paste0('pvalue_',BL))]
  int_p <- int_p[match(pheno_list,int_p$neighbour_phenotype,),paste0('pvalue_',BL)]
  rowMeans(int_p,na.rm = T)
})
rownames(int_pval) <- pheno_list
interactions_signif <- gtools::stars.pval(int_pval)

pheatmap(int_avg,scale = 'none',na_col = "white",display_numbers =interactions_signif, cluster_rows = F,main = "All samples: Averages BL",
         fontsize = 20,cluster_cols = F,cellwidth = 25,cellheight = 25,color = hcl.colors(palette = "Blue-Red 2",200),angle_col = 45,breaks = seq(-1,1,0.01))

```

# Interactions Carbo C2
```{r,fig.height=16,fig.width=16}
library(pheatmap)
#extract spatial interactions from object
pheno_list <- setdiff(unique(interactions$phenotype),c("all_neg","Autofluo","other"))
C2_Carbo <- unique(sobj$orig.ident[sobj$timepoint_s != "BL" & sobj$Tx=="Carboplatin"])

int_avg <- sapply(pheno_list,function(pop){
  int_df <- interactions[interactions$phenotype==pop,c("phenotype","neighbour_phenotype",C2_Carbo)]
  int_df <- int_df[match(pheno_list,int_df$neighbour_phenotype,),C2_Carbo]
  rowMeans(int_df,na.rm = T)
})
rownames(int_avg) <- pheno_list

#summarize spatial interactions for all images
int_pval <- sapply(pheno_list,function(pop){
  int_p <- interactions[interactions$phenotype==pop,c("phenotype","neighbour_phenotype",paste0('pvalue_',C2_Carbo))]
  int_p <- int_p[match(pheno_list,int_p$neighbour_phenotype,),paste0('pvalue_',C2_Carbo)]
  rowMeans(int_p,na.rm = T)
})
rownames(int_pval) <- pheno_list
interactions_signif <- gtools::stars.pval(int_pval)

pheatmap(int_avg,scale = 'none',na_col = "white",display_numbers =interactions_signif, cluster_rows = F,main = "Carboplatin: Averages C2",
         fontsize = 20,cluster_cols = F,cellwidth = 25,cellheight = 25,color = hcl.colors(palette = "Blue-Red 2",200),angle_col = 45,breaks = seq(-1,1,0.01))

```

## Annotation colors
```{r}
rep_colors <- hcl.colors(palette = 'Temps',5,rev=T)[c(1,3,4,5)]
ann_colors = list(
  Tx= c('Carboplatin'="#FDC086",
        "Carboplatin + Nivolumab"="#386CB0"),
  Response = c(CR=rep_colors[4],
               PR=rep_colors[3],
               SD=rep_colors[2],
               PD=rep_colors[1]),
  Evaluation=c(Excellent=rep_colors[4],Poor=rep_colors[1]),
  timepoint=c(BL="white",
              C2='lightblue',
              C3="royalblue1",
              EOT="darkblue"))
```

## Visualize all interactions
```{r,fig.height=8,fig.width=25}
pheno_list <- setdiff(unique(interactions$phenotype),c("all_neg","Autofluo","other"))
all_sample<- unique(sobj$orig.ident)
interactions <- interactions[interactions$phenotype %in% pheno_list ,]
interactions <- interactions[interactions$neighbour_phenotype %in% pheno_list ,]

#Order : 
# timepoint / TX / Evaluation 
BL_C_E <- unique(sobj$orig.ident[sobj$timepoint_s == "BL" & sobj$Tx == "Carboplatin" & sobj$Evaluation=="Excellent"])
BL_C_P <- unique(sobj$orig.ident[sobj$timepoint_s == "BL" & sobj$Tx == "Carboplatin" & sobj$Evaluation=="Poor"])

C2_C_E <- unique(sobj$orig.ident[sobj$timepoint_s != "BL" & sobj$Tx == "Carboplatin" & sobj$Evaluation=="Excellent"])
C2_C_P <- unique(sobj$orig.ident[sobj$timepoint_s != "BL" & sobj$Tx == "Carboplatin" & sobj$Evaluation=="Poor"])

BL_CN_E <- unique(sobj$orig.ident[sobj$timepoint_s == "BL" & sobj$Tx == "Carboplatin + Nivolumab" & sobj$Evaluation=="Excellent"])
BL_CN_P <- unique(sobj$orig.ident[sobj$timepoint_s == "BL" & sobj$Tx == "Carboplatin + Nivolumab" & sobj$Evaluation=="Poor"])

C2_CN_E <- unique(sobj$orig.ident[sobj$timepoint_s != "BL" & sobj$Tx == "Carboplatin + Nivolumab" & sobj$Evaluation=="Excellent"])
C2_CN_P <- unique(sobj$orig.ident[sobj$timepoint_s != "BL" & sobj$Tx == "Carboplatin + Nivolumab" & sobj$Evaluation=="Poor"])

int_all <- interactions[,c(BL_C_P,BL_CN_P,
                           BL_C_E,BL_CN_E,
                           C2_C_P,C2_CN_P,
                           C2_C_E,C2_CN_E)]
rownames(int_all) <- paste0(interactions$phenotype, "->",interactions$neighbour_phenotype)

#remove self interaction
int_all <- int_all[interactions$phenotype != interactions$neighbour_phenotype,]
ann_pheatmap <- sapply(all_sample, function(x) {
  rowdf <- sobj@meta.data[sobj@meta.data$orig.ident == x, c('Tx', 'Evaluation', 'Response', 'timepoint')][1, ]
  # Converting factor columns to character type
  rowdf$Evaluation <- as.character(rowdf$Evaluation)
  rowdf$Response <- as.character(rowdf$Response)
  return(unlist(rowdf))
})

# Transposing and then converting to a dataframe with stringsAsFactors set to FALSE
ann_pheatmap <- as.data.frame(t(as.data.frame(ann_pheatmap)), stringsAsFactors = FALSE)

pheatmap(t(int_all),scale = 'none',na_col = "white",
         #display_numbers =interactions_signif, 
         cluster_rows = F,cluster_cols = F,
         main = "All samples: Averages",
         annotation_row = ann_pheatmap,annotation_colors = ann_colors,
         fontsize = 5,cellwidth = 5,cellheight = 10,
         color = pals::coolwarm(200),
         angle_col = 90,gaps_row = seq(0,39,5))
```

## Distances 
```{r,fig.height=7,fig.width=7}
average_dists <- sapply(pheno_list,function(pop){
  print(pop)
  ids <- which(sobj$annotation==pop)
  mns <- colMeans(sobj@misc$dists[ids,],na.rm=T)
})
colnames(average_dists) <- rownames(average_dists)
average_dists <- average_dists[pheno_list,pheno_list]
#average_dists <- scale(as.data.frame(average_dists))
pheatmap(average_dists,scale = "row",cluster_rows = F,cluster_cols = F,cellwidth = 20,cellheight = 20,na_col = "white",color = rev(pals::coolwarm(100)),angle_col = 45)
```
